---
title: "Three (Groups of) Blind Mice" 
subtitle: "Familial Clusters of Cataract Development in Irradiated Mice"
author: "Alyssa Allsop and Amira Burns"
date: "`r Sys.Date()`"
output: 
  html_document: 
    fig_caption: true
    
  pdf_document: true
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # change to echo = FALSE for final
```

```{r load_libs}
library(readxl)
library(tidyverse)
library(lme4)
library(broom.mixed)
library(kableExtra)
library(ggsci)
library(sjPlot)
library(coda)
library(rjags)
library(R2jags)
library(superdiag)
library(mcmcplots)
library(ggmcmc)
```

```{r load_data}
cats <- read_excel("GRSD.cataract.xlsx", sheet = "Sheet1")
```

```{r format_data}
# remove spaces from column and value names
names(cats) <- str_replace_all(names(cats), " ", "_")
cats <- cats %>%
  mutate(CoatColor = str_replace_all(coat_color, " ", "_"))
str(cats)

# turn categorical vars into factor
cats <- cats %>%
  rename(Age = `age_(days)`,
         Weight = weight,
         Animal = animal) %>%
  mutate(Sex = as.factor(sex),
         CoatColor = as.factor(CoatColor),
         Family = as.factor(family), # should this stay a factor? Yes?
         BCS = as.ordered(BCS),
         Treatment = relevel(as.factor(groups), ref = "Unirradiated"),
         MyeloidLeukemia = as.factor(Myeloid_Leukemia),
         HarderianTumor = as.factor(Harderian_Tumor),
         PreTLymphoma = as.factor(PreT_Lymphoma),
         Score = as.ordered(Cataract_Score)) # ordinal cat; leave as numeric?

# select vars, add binary conversion for score
cats <- cats %>%
  select(c(Animal, Sex, Weight, CoatColor, Family, BCS, Age, Treatment,
           MyeloidLeukemia, HarderianTumor, PreTLymphoma, Score)) %>%
  mutate(Cataracts = ifelse(Score < 2, 0, 1))

# glance at the dataset
# str(cats)
```

## Abstract

## Introduction  
Little is known about the effects of high atomic number and energy (HZE) radiation, a main component of space radiation to which exposure is unavoidable beyond Earth’s magnetic field. In contrast, extensive research shows exposure to high doses of gamma radiation leads to acute radiation sickness. Effects include damage to the blood forming system, GI system, immune system, increased risk for cancer, cardiovascular disease, neurodegenerative disease, death, and cataracts.. The serious health implications from HZE radiation for astronauts who leave Earth’s magnetic field warrant further study. Adverse effects from radiation may also be attributable to other factors, including genetics; several different cancers have been observed to cluster in mice families (Chernyavskiy, et. al., 2017). Accounting for potential family clustering allows for thorough examination of the primary research questions. Is there a genetic susceptibility to cataracts caused by radiation? Accounting for potential genetic susceptibility, is there a difference in cataract presentation between HZE radiation and gamma radiation?
The full dataset includes 1820 unique mice from 48 unique families, with equal random assignments by family to each of three treatment groups. Mice are bred over several generations to create a genetically heterogeneous sample, the better to represent the diverse biology of the human population. The treatments are HZE irradiation, gamma irradiation, and non-irradiated control. The HZE group is irradiated with either silicon or iron nuclei HZE ions, which are considered as a single treatment group. The second group is subjected to 137 Cs gamma irradiation. Mice in both irradiated groups are exposed to radiation at 7-12 weeks of age. The third group is unirradiated control. All mice are monitored until 800 days of age –effectively a survival study. Mice are checked weekly for symptoms of cataracts, cataract risk factors, and other symptoms of radiation exposure such as tumors and carcinomas. Resources do not allow for weekly measurement of every mouse; previous measurements are carried forward if a mouse is not assessed on a particular week. 
This analysis uses a simplified version of the data set - a snapshot of the 1169 mice that are alive at 552 days. This cutoff is chosen because it is the median survival time for the group with the shortest median survival. There are 47 unique families with n = 396 in the HZE group, n = 277 in the gamma radiation group, and n = 496 in the unirradiated control group. There are up to two generations of mice pups from each family in the dataset, but generation is not distinguished in the simplified data. Family size ranges from n = 11 to n = 48, with median family size of 24.


## Summary Statistics  

```{r table}
sex_trt <- cats %>%
  group_by(Treatment, Sex, Cataracts) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  group_by(Treatment, Sex) %>%
  mutate(prop = round(n/sum(n), digits = 2))

sum_tab <- sex_trt %>%
  pivot_wider(names_from = Sex, values_from = c(n, prop)) %>%
  mutate(Cataracts = ifelse(Cataracts == 0, "No", "Yes"),
         Treatment = rep(" ", 1)) %>%
  relocate(prop_F, .after = n_F)
n_C <- c(438, 58, 214, 63, 281, 115)
prop_C <- c(0.88, 0.12, 0.77, 0.23, 0.71, 0.29)

sum_tab <- sum_tab %>%
  ungroup() %>%
  mutate(n_C = n_C, prop_C = prop_C) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 258, prop_F = NA,
          n_M = 238, prop_M = NA, n_C = 496, prop_C = NA, .after = 2) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 145, prop_F = NA,
          n_M = 132, prop_M = NA, n_C = 277, prop_C = NA, .after = 5) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 205, prop_F = NA,
          n_M = 191, prop_M = NA, n_C = 396, prop_C = NA, .after = 8) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 608, prop_F = NA,
          n_M = 561, prop_M = NA, n_C = 1169, prop_C = NA, .after = 9)

options(knitr.kable.NA = '')
ct_prop <- kbl(sum_tab,
               caption = "Mice Counts and Percentages by Sex, Treatment Group, Cataracts",
               col.names = c("Treatment", "Cataracts", "n", "prop", "n", "prop", "n", "prop")) %>%
  kable_styling(latex_options = "hold_position") %>%
  kable_classic_2(full_width = F, html_font = "Cambria") %>%
  add_header_above(c(" " = 2, "Female" = 2, "Male" = 2, "$\\sum$" = 2)) %>%
  pack_rows(index = c("Control" = 3, "Gamma" = 3, "HZE" = 3, "$\\sum$" = 1))
ct_prop
```

```{r lineplot}
# Plot of average score of sex-group-family
grsex_score <- cats %>%
  group_by(Sex, Treatment, Family) %>%
  summarize(mean_score = mean(Cataracts))

# choose facet direction and just show one
ggplot(grsex_score, aes(x = Treatment, y = mean_score, color = Sex)) +
  geom_line(aes(group = Family)) + facet_grid(vars(Sex)) +
  scale_color_startrek() +
  scale_x_discrete(expand = c(0, .2)) +
  theme_light() +
  labs(y = "mean score",
       title = "Cataracts by Family, Sex, Treatment Group")

ggplot(grsex_score, aes(x = Treatment, y = mean_score, color = Sex)) +
  geom_line(aes(group = Family)) + facet_wrap(vars(Sex)) +
  scale_color_startrek() +
  scale_x_discrete(expand = c(0, .2)) +
  theme_light() +
  labs(y = "mean score",
       title = "Cataracts by Family, Sex, Treatment Group")
```

```{r barplot}
# barplot of proportion of each group with cataracts
sex_trt <- cats %>%
  group_by(Treatment, Sex, Cataracts) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  group_by(Treatment, Sex) %>%
  mutate(prop = round(n/sum(n), digits = 2))

cats_grp <- sex_trt %>%
  filter(Cataracts == 1)

ggplot(cats_grp, aes(x = Sex, y = prop, fill = Treatment)) +
  geom_col(position = "dodge") +
  scale_fill_startrek() +
  theme_minimal() +
  ggtitle("Sample Proportion with Cataracts by Sex and Treatment Group")

```


## Statistical Methods

### Frequentist Approach
```{r glmm}
# mixed model binomial logistic regression using adaptive Gauss-Hermite quadrature (nAGQ = 25)

mod <- glmer(Cataracts ~ 0 + Treatment*Sex + (1|Family), data = cats, family = binomial)

summary(mod)
```

```{r glmm_ors}
plot_model(mod, sort.est = TRUE, show.values = TRUE, type = "int",
           color = "Dark2", vline.color = "darkorchid3",
           width = 0.1, title = "Final Model: Cataracts Odds Ratios by Sex, Treatment Group")
tab_model(mod, show.re.var = TRUE,
          pred.labels = c("Control(Female)", "Gamma(Female)", "HZE(Female)",
                          "Control(Male)", "Gamma(Male)", "HZE(Male)"),
          dv.labels = "Final Model Effects of Treatment on Cataracts")
```
### Bayesian Approach  
```{r spec_mod}
# Specify the model
cat("model{
  for(i in 1:N){
    CAT[i] ~ dbern(p[i])     # Bernoulli-distributed response
    logit(p[i])<- b0*Unirradiated[i] + b1*Gamma[i] + b2*HZE[i] + b3*Male[i] +
    b4*Male[i]*Gamma[i] + b5*Male[i]*HZE[i] + a[Family[i]]   # likelihood function
  }
  for(j in 1:nFam){
    a[j] ~ dnorm(0, tau)
  }
  b0 ~ dnorm(0.0, 1.0E-4)   # vaguely informative priors
  b1 ~ dnorm(0.0, 1.0E-4)
  b2 ~ dnorm(0.0, 1.0E-4)
  b3 ~ dnorm(0.0, 1.0E-4)
  b4 ~ dnorm(0.0, 1.0E-4)
  b5 ~ dnorm(0.0, 1.0E-4)
  tau ~ dgamma(1.0E-3,1.0E-3)
  sigma2 <- 1/tau     # convert precision 'tau' to variance 'sigma2'
}", file = "cat.jag")
```

```{r prep_dat}
# Prepare the data for JAGS
# break Treatment into dummy variables for each group
treatment <- model.matrix(~ Treatment - 1, cats)
sex <- model.matrix(~Sex -1, cats)
colnames(treatment) <- c("Unirradiated", "Gamma", "HZE")
cats <- data.frame(cats, treatment, sex)

# format relevant data as a list
data <- list(CAT = cats$Cataracts, Unirradiated = cats$Unirradiated, Gamma = cats$Gamma, 
             HZE = cats$HZE, Male = cats$SexM, Family = cats$Family, 
             nFam = length(unique(cats$Family)), N = nrow(cats))
```

```{r setup_alg}
nIter <- 60000
nChains <- 3
nThin <- 1
BurnIn <- 10000
nAdapt <- 1000
ests <- summary(mod)$coef[,1] # pull starting values from frequentist model
var <- as.numeric(as.data.frame(VarCorr(mod))$vcov)
inits <- list(list("tau" = var+0.2, "b0" = ests[1]+0.5, "b1" = ests[2]+0.5, "b2" = ests[3]+0.5,
                   "b3" = ests[4]+0.2, "b4" = ests[5]+0.2, "b5" = ests[6]+0.2),
              list("tau" = var-0.2, "b0" = ests[1]-0.5, "b1" = ests[2]-0.5, "b2" = ests[3]-0.5,
                   "b3" = ests[4]-0.2, "b4" = ests[5]-0.2, "b5" = ests[6]-0.2),
              list("tau" = var, "b0" = ests[1], "b1" = ests[2], "b2" = ests[3],
                   "b3" = ests[4], "b4" = ests[5], "b5" = ests[6]))
```

```{r run_bayes}
# -- Compile and run the model
params <- c("b0", "b1", "b2", "b3", "b4", "b5", "sigma2")
set.seed(556)
model.fit <- jags(data = data,
                  inits = inits,
                  parameters.to.save = params,
                  model.file = "cat.jag",
                  n.chains = nChains,
                  n.iter = nIter,
                  n.burnin = BurnIn,
                  n.thin = nThin)

```

```{r bayes_summary}
mcmc.model <- as.mcmc(model.fit)
summary(mcmc.model)

# extract mean estimates
est_means <- summary(mcmc.model)$statistics[,1]

# calculate highest posterior density (HPD) credible intervals
hpds <- hpds <- HPDinterval(mcmc.model[[3]])
```

## Limitations and Alternatives

Presence of competing risks in the experimental design obscure the presence of family clusters of cataracts. It is reasonable to assume that individuals in the irradiated Treatment groups were more likely to die due to more acute effects of irradiation before developing cataracts, compared to individuals in the control group. Simplifying the dataset by instituting a cutoff age attempts to address this limitation, but the issue persists even after excluding individuals deceased before Age = 552.  


## Results and Conclusions

## Authors' Statements  
  
Both authors contributed to statistical analysis and writing for this report. Both authors read and approved the final report. Both authors contributed equally to introduction, exploratory data analysis and summary statistics. Alyssa Allsop served as the primary statistician on Frequentist models; Amira Burns served as the primary statistician on Bayesian models. 

## References  

Agresti, A (2013). *Categorical Data Analysis* (Third Edition). John Wiley & Sons.  
  
Chernyavskiy, P., Edmondson, E. F., Weil, M. M., & Little, M. P. (2017). *High-energy particle beam and gamma radiation exposure, familial relatedness and cancer in mice.* British journal of cancer, 117(1), 41–50. https://doi.org/10.1038/bjc.2017.141  
  
Gelman, A., Carlin, J., Stern, H., Dunson, D., Vehtari, A., Rubin, D. (2014). *Bayesian Data Analysis* (Third Edition). CRC Press.  
  
Roback, P., Legler, J. (2021). *Beyond Multiple Linear Regression. Applied Generalized Linear Models and Multilevel Models in R* (Second Edition). CRC Press. Online version: https://bookdown.org/roback/bookdown-BeyondMLR/  
  
DiMaggio, C., Albert, J., Spiegelhalter, D., Best, N., Gelman, A., Carstensen, B., Guerrin, L., Jensen, S. (2015) *Bayesian Analysis for Epidemiologists III: Regression Modeling.* Center for Injury Epidemiology and Prevention at Columbia University.  http://www.columbia.edu/~cjd11/charles_dimaggio/DIRE/styled-4/styled-11/code-8/#bayesian-regression-modeling-with-bugs-or-jags

## Appendix
