---
title: "Three (Groups of) Blind Mice" 
subtitle: "Familial Clusters of Cataract Development in Irradiated Mice"
author: "Alyssa Allsop and Amira Burns"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
** data prep to allow this to be a stand alone script**
```{r}
library(readxl)
library(tidyverse)
library(lmerTest)
library(broom.mixed)
library(kableExtra)
library(ggsci)

cats <- read_excel("GRSD.cataract.xlsx", sheet = "Sheet1")
# remove spaces from column and value names
names(cats) <- str_replace_all(names(cats), " ", "_")
cats <- cats %>%
  mutate(coat_color = str_replace_all(coat_color, " ", "_"))
str(cats)

# turn categorical vars into factor
cats <- cats %>%
  rename(age = `age_(days)`) %>%
  mutate(sex = as.factor(sex),
         coat_color = as.factor(coat_color),
         fam = as.factor(family), # should this stay a factor? Yes?
         BCS = as.ordered(BCS),
         group = as.factor(group),
         groups = as.factor(groups),
         myeloid_leuk = as.factor(Myeloid_Leukemia),
         harderian_tumor = as.factor(Harderian_Tumor),
         preT_lymph = as.factor(PreT_Lymphoma),
         cat_score = as.ordered(Cataract_Score))
# add column for binary response
cats <- cats %>%
  mutate(cataract_status = ifelse(Cataract_Score >= 2, 1, 0))
```


## Abstract

## Introduction

## Summary Statistics

```{r}
#table showing proportions of cataracts (>=2) for groups and sex
xtabs(cataract_status ~ groups + sex, cats)/1169 #divide to get proportions

#barplot also showing proportions of cataracts(>= 2) for groups and sex
barplot(xtabs(cataract_status ~ groups + sex, cats)/1169, beside = TRUE, legend.text = TRUE)

```

## Statistical Methods

```{r}
# full mod mixed model binomial logistic regression using adaptive Gauss-Hermite quadrature (nAGQ = 25)

binary_mod_full <- glmer(cataract_status ~ sex + weight + BCS +
                     groups + Myeloid_Leukemia + Harderian_Tumor +
                     PreT_Lymphoma + coat_color + (1|family), nAGQ = 25,
                    family = binomial, data = cats)
summary(binary_mod_full)


#reduced model (removing variables that aren't highly significant)
binary_mod_red <- glmer(cataract_status ~ sex + weight + groups + 
                          (1|family), nAGQ = 25, family = binomial, data = cats)
summary(binary_mod_red)

# use ANOVA to compare the two models
anova(binary_mod_full, binary_mod_red) 
#high p value indicates the reduced model is better. The reduced model also has a smaller AIC
```

## Limitations and Alternatives

Using an ANOVA to compare glmer models uses standard likelihood-based methods to construct a chi-squared test so we must view those results with caution.

## Results and Conclusions

## Authors' Statements

## References

## Appendix
