---
title: "Three (Groups of) Blind Mice" 
subtitle: "Familial Clusters of Cataract Development in Irradiated Mice"
author: "Alyssa Allsop and Amira Burns"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE) # change to echo = FALSE for final
```

```{r load_libs, warning = FALSE, message = FALSE}
library(readxl)
library(tidyverse)
library(lme4)
library(broom.mixed)
library(kableExtra)
library(emmeans)
library(ggsci)
library(sjPlot)
library(coda)
library(rjags)
library(R2jags)
library(superdiag)
library(mcmcplots)
library(ggmcmc)
library(gridExtra)
```

```{r load_data}
cats <- read_excel("GRSD.cataract.xlsx", sheet = "Sheet1")
```

```{r format_data}
# remove spaces from column and value names
names(cats) <- str_replace_all(names(cats), " ", "_")
cats <- cats %>%
  mutate(CoatColor = str_replace_all(coat_color, " ", "_"))

# turn categorical vars into factor
cats <- cats %>%
  rename(Age = `age_(days)`,
         Weight = weight,
         Animal = animal) %>%
  mutate(Sex = as.factor(sex),
         CoatColor = as.factor(CoatColor),
         Family = as.factor(family), # should this stay a factor? Yes?
         BCS = as.ordered(BCS),
         Treatment = relevel(as.factor(groups), ref = "Unirradiated"),
         MyeloidLeukemia = as.factor(Myeloid_Leukemia),
         HarderianTumor = as.factor(Harderian_Tumor),
         PreTLymphoma = as.factor(PreT_Lymphoma),
         Score = as.ordered(Cataract_Score)) # ordinal cat; leave as numeric?

# select vars, add binary conversion for score
cats <- cats %>%
  select(c(Animal, Sex, Weight, CoatColor, Family, BCS, Age, Treatment,
           MyeloidLeukemia, HarderianTumor, PreTLymphoma, Score)) %>%
  mutate(Cataracts = ifelse(Score < 2, 0, 1))

# glance at the dataset
# str(cats)
```

## Abstract

As astronauts travel farther into space, outside of the earth’s protective magnetic field, they are exposed to radiation that could have serious effects. One of those side effects is cataract development. It has also been thought that there might be a genetic predisposition to develop cataracts that is passed down in families. The goal of this study was to use mice in families exposed to radiation to determine if there is cataract clustering in families. 
The mice were bread in heterogeneous diverse families to try to mimic the diversity of human genes. Within families, mice were assigned to receive either HZE irradiation, gamma irradiation, or no radiation. From there, the mice were monitored and symptoms recorded at time of death for mice that lived to at least 552 days. 
A mixed effects model was created to account for the family effect on cataract development. The model included terms for sex, treatment group, sex by treatment group interaction, and a random intercept for family. Family was shown to be an important source of variation in predicting cataract development. Males had big differences in cataract development among treatment groups whereas females had only small differences. Males were also more likely to develop severe cataracts than females in all three treatment groups.
There are other risks involved in radiation exposure that could cause death before cataracts have time to develop. This complicates the inferences made about cataract development from this study alone. Overall, this is a great starting point into studying how humans might react to radiation as they leave the earth’s magnetic field


## Introduction

Little is known about the effects of high atomic number and energy (HZE) radiation, a main component of space radiation to which exposure is unavoidable beyond Earth's magnetic field. In contrast, extensive research shows exposure to high doses of gamma radiation leads to acute radiation sickness. Effects include damage to the blood forming system, GI system, immune system, increased risk for cancer, cardiovascular disease, neurodegenerative disease, death, and cataracts.. The serious health implications from HZE radiation for astronauts who leave Earth's magnetic field warrant further study. Adverse effects from radiation may also be attributable to other factors, including genetics; several different cancers have been observed to cluster in mice families (Chernyavskiy, et. al., 2017). Accounting for potential family clustering allows for thorough examination of the primary research questions. Is there a genetic susceptibility to cataracts caused by radiation? Accounting for potential genetic susceptibility, is there a difference in cataract presentation between HZE radiation and gamma radiation? The full dataset includes 1820 unique mice from 48 unique families, with equal random assignments by family to each of three treatment groups. Mice are bred over several generations to create a genetically heterogeneous sample, the better to represent the diverse biology of the human population. The treatments are HZE irradiation, gamma irradiation, and non-irradiated control. The HZE group is irradiated with either silicon or iron nuclei HZE ions, which are considered as a single treatment group. The second group is subjected to 137 Cs gamma irradiation. Mice in both irradiated groups are exposed to radiation at 7-12 weeks of age. The third group is unirradiated control. All mice are monitored until 800 days of age --effectively a survival study. Mice are checked weekly for symptoms of cataracts, cataract risk factors, and other symptoms of radiation exposure such as tumors and carcinomas. Resources do not allow for weekly measurement of every mouse; previous measurements are carried forward if a mouse is not assessed on a particular week. This analysis uses a simplified version of the data set - a snapshot of the 1169 mice that are alive at 552 days. This cutoff is chosen because it is the median survival time for the group with the shortest median survival. There are 47 unique families with n = 396 in the HZE group, n = 277 in the gamma radiation group, and n = 496 in the unirradiated control group. There are up to two generations of mice pups from each family in the dataset, but generation is not distinguished in the simplified data. Family size ranges from n = 11 to n = 48, with median family size of 24.

```{r scoregrps}
cats %>% group_by(Score) %>%
  count(Treatment) %>%
  pivot_wider(names_from = Score, values_from = n) %>%
  group_by(Treatment) %>%
  mutate(Total = sum(c(`1`, `2`, `3`, `4`))) %>%
  kbl(caption = "Counts of Score by Treatment") %>%
  add_header_above(c( " " = 1, "Cataract Score" = 4, " " = 1)) %>%
  kable_classic(full_width = F, html_font = "Cambria", font_size = 12, position = "float_right") 
```

The response is Merriam-Focht cataract score (Merriam & Focht, 1957), an ordinal categorical variable corresponding to radiation-associated ocular changes in the eye. A score of 0 is associated with a completely clear lens, while a score of 5 is associated with a completely occluded lens. This dataset contains cataract score levels = [1, 2, 3, 4]. A score $\ge$ 2 indicates presence of cataracts, and the small sample sizes for score $>$ 2 across all treatment groups raise concerns about making inference on an ordinal analysis; consequently, the response is converted to binary *Cataracts* with score = 1 converted to 0, and score $\ge$ 2 converted to 1. The main experimental factor is *Treatment*, a categorical variable with three levels = [HZE radiation, gamma radiation, non-irradiated control]. A single random effect, genetic *Family*, is a categorical factor with 47 levels. Both Treatment and Family are central to the experimental design and will serve as the basis for all models considered. Additional covariates under consideration are:\
+ Sex: categorical with two levels = [Male, Female].\
+ Weight: numeric, in grams; range = [16.11, 59.85].\
+ Coat color: categorical with six levels = [albino, black, creme brown, grey, light brown, light grey].\
+ Body condition score (BCS): ordered categorical with 5 levels, representing a range from emaciated to obese. Levels = [BC1, BC2, BC3, BC4, BC5].\
+ Age: numeric, in days; range = [552, 816].\
+ Three categorical variables indicating presence of other cancers: Harderian Tumor: categorical with three levels = [bilateral, unilateral, none]. Myeloid Leukemia: binary categorical with two levels = [0, 1]. PreT Lymphoma: binary categorical with two levels = [0, 1].  
  
The experimental design prescribed the inclusion of Treatment and Family in a final model. 


## Summary Statistics

Exploratory data analysis consisted of investigating the response in terms of Treatment and Family and any potential associations between the response and covariates. In addition, possible interactions and correlations between covariates and the primary predictors in terms of the response were of interest.

```{r table, message = FALSE}
sex_trt <- cats %>%
  group_by(Treatment, Sex, Cataracts) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  group_by(Treatment, Sex) %>%
  mutate(prop = round(n/sum(n), digits = 2))

sum_tab <- sex_trt %>%
  pivot_wider(names_from = Sex, values_from = c(n, prop)) %>%
  mutate(Cataracts = ifelse(Cataracts == 0, "No", "Yes"),
         Treatment = rep(" ", 1)) %>%
  relocate(prop_F, .after = n_F)
n_C <- c(438, 58, 214, 63, 281, 115)
prop_C <- c(0.88, 0.12, 0.77, 0.23, 0.71, 0.29)

sum_tab <- sum_tab %>%
  ungroup() %>%
  mutate(n_C = n_C, prop_C = prop_C) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 258, prop_F = NA,
          n_M = 238, prop_M = NA, n_C = 496, prop_C = NA, .after = 2) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 145, prop_F = NA,
          n_M = 132, prop_M = NA, n_C = 277, prop_C = NA, .after = 5) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 205, prop_F = NA,
          n_M = 191, prop_M = NA, n_C = 396, prop_C = NA, .after = 8) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 608, prop_F = NA,
          n_M = 561, prop_M = NA, n_C = 1169, prop_C = NA, .after = 9)

options(knitr.kable.NA = '')
ct_prop <- kbl(sum_tab,
               caption = "Mice Counts and Percentages by Sex, Treatment Group, Cataracts",
               col.names = c("Treatment", "Cataracts", "n", "prop", "n", "prop", "n", "prop")) %>%
  kable_styling(latex_options = "hold_position") %>%
  kable_classic_2(full_width = F, html_font = "Cambria") %>%
  add_header_above(c(" " = 2, "Female" = 2, "Male" = 2, "$\\sum$" = 2)) %>%
  pack_rows(index = c("Control" = 3, "Gamma" = 3, "HZE" = 3, "$\\sum$" = 1))
ct_prop
```

```{r lineplot, message = FALSE}
# Plot of average score of sex-group-family
grsex_score <- cats %>%
  group_by(Sex, Treatment, Family) %>%
  summarize(mean_score = mean(Cataracts))

# choose facet direction and just show one
l1 <- ggplot(grsex_score, aes(x = Treatment, y = mean_score, color = Sex)) +
  geom_line(aes(group = Family)) + facet_grid(vars(Sex)) +
  scale_color_startrek() +
  scale_x_discrete(expand = c(0, .2)) +
  theme_light() +
  labs(y = "mean score",
       title = "Cataracts by Family, Sex, Treatment Group")
```

```{r barplot, message = FALSE}
# barplot of proportion of each group with cataracts
sex_trt <- cats %>%
  group_by(Treatment, Sex, Cataracts) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  group_by(Treatment, Sex) %>%
  mutate(prop = round(n/sum(n), digits = 2))

cats_grp <- sex_trt %>%
  filter(Cataracts == 1)

b1 <- ggplot(cats_grp, aes(x = Sex, y = prop, fill = Treatment)) +
  geom_col(position = "dodge") +
  scale_fill_startrek() +
  theme_light() +
  ggtitle("Sample Proportion")

logits <- sex_trt %>%
  group_by(Treatment, Sex) %>%
  summarise(odds_ratio = prop[Cataracts == 1]/prop[Cataracts == 0])

b2 <- ggplot(logits, aes(x = Sex, y = odds_ratio, fill = Treatment)) +
  geom_col(position = "dodge") +
  scale_fill_startrek() +
  theme_light() +
  labs(y = "Odds Ratio",
       title = "Empirical Odds Ratios")

grid.arrange(b1, b2, ncol = 2, top = "Cataracts by Sex, Treatment")
```

## Statistical Methods

### Frequentist Approach

```{r glmm}
# mixed model binomial logistic regression 

full_mod <- glmer(Cataracts ~ Treatment + Sex + Weight + CoatColor + BCS + HarderianTumor +
                    MyeloidLeukemia + PreTLymphoma + (1|Family), data = cats, family = binomial)
fixed_mod <- glm(Cataracts ~ Treatment*Sex, data = cats, family = binomial)
no_interaction <- glmer(Cataracts ~ Treatment + Sex + (1|Family), data = cats, family = binomial)
final_mod <- glmer(Cataracts ~ Treatment*Sex + (1|Family), data = cats, family = binomial)
summary(final_mod)
```


```{r probs}
# convert estimates to probabilities
ests <- summary(final_mod)$coef[,1]
ps <- exp(ests)/(1+exp(ests))
```

In order to study the family/genetic effect on cataract development among mice, a mixed model with a random intercept for family was fit. From plot ______, it was apparent that there was an interaction between sex and treatment group and therefore an interaction term was included in the model. The final model was a generalized mixed logistic regression model after converting the response to a binary 0, 1. The terms included in the final model were: sex, treatment group, sex by treatment group interaction, and a random intercept for family. The full model is written as follows:

$Y_{ij} \sim Bernoulli(p)$  

$$\begin{aligned}
log(\frac{p}{1-p}) = &\ \beta_0\ +\beta_1*Gamma_i\ + \beta_2*HZE_i\ + \beta_3*M_i\ + \\ &\beta_4*Gamma_i*M_i\ + \beta_5*HZE_i*M_i\ +\ [v_{j} + \epsilon_{ij}] \\
&i = 1, ..., 1169\ \mbox{ mice} \\
&j = 1,...,47\ \ \mbox{ families}
\end{aligned}$$

Where M represents males, v is the variance of the random intercept for family, and $\epsilon$ is the error variance.

Akaike's Information Criterion (AIC) was used in the model selection process. Other models considered included: a mixed model with all predictors included in the dataset, a mixed model with only treatment and sex and no interaction, a fixed effects model with sex, treatment, and sex by treatment interaction. 



### Bayesian Approach

We fit a complementary Bayesian model to examine the distribution of the estimates from our final model, as well as test its robustness to alternative approaches. The Bayes model is parameterized as:

$$
\begin{aligned}
Y_{ij} \sim &Bernoulli(p)\\
log(\frac{p}{1-p}) = &\ \beta_0*Control_i*F_i\ +\beta_1*Gamma_i*F_i\ + \beta_2*HZE_i*F_i\ + \\ &\beta_3*Control_i*M_i\ +\beta_4*Gamma*M\ + \beta_5*HZE*M\ + \\
&v_{j} + \epsilon_{ij}\\
&i = 1, ..., 1169\ \mbox{ mice} \\
&j = 1,...,47\ \ \mbox{ families, and} \\
v_j \sim\ &N(0, \tau)
\end{aligned}
$$

With non-informative parameters:\
$$
\begin{aligned}
\beta_0 &\sim N(0, 0.001)\\
\beta_1 &\sim N(0, 0.001)\\
\beta_2 &\sim N(0, 0.001)\\
\beta_3 &\sim N(0, 0.001)\\
\beta_4 &\sim N(0, 0.001)\\
\beta_5 &\sim N(0, 0.001)\\
v_i &\sim N(0, \sigma^2)\\
\tau &\sim Gamma(0.001, 0.001) \mbox{ where}\ \sigma^2 = 1/\tau\\
\end{aligned}
$$

The model was run through a Gibbs sampler Markov Chain Monte Carlo (MCMC) algorithm to approximate the posterior distribution of all fixed effects estimates and the variance of the random effect. 3 chains of 60000 total iterations with a 10000-iteration burn-in period; starting values were obtained from the estimates generated by the Frequentist model, with noise added to avoid false convergence.

```{r bayes_mod, message = FALSE, cache=TRUE}
# Specify the model
cat("model{
  for(i in 1:N){
    CAT[i] ~ dbern(p[i])     # Bernoulli-distributed response
    logit(p[i])<- b0+ b1*Gamma[i] + b2*HZE[i] + b3*Male[i] +
    b4*Male[i]*Gamma[i] + b5*Male[i]*HZE[i] + a[Family[i]]   # likelihood function
  }
  for(j in 1:nFam){
    a[j] ~ dnorm(0, tau)
  }
  b0 ~ dnorm(0.0, 1.0E-3)   # vaguely informative priors
  b1 ~ dnorm(0.0, 1.0E-3)
  b2 ~ dnorm(0.0, 1.0E-3)
  b3 ~ dnorm(0.0, 1.0E-3)
  b4 ~ dnorm(0.0, 1.0E-3)
  b5 ~ dnorm(0.0, 1.0E-3)
  tau ~ dgamma(1.0E-3,1.0E-3)
  sigma2 <- 1/tau     # convert precision 'tau' to variance 'sigma2'
}", file = "cat.jag")

# Prepare the data for JAGS
# break Treatment into dummy variables for each group
treatment <- model.matrix(~ Treatment - 1, cats)
sex <- model.matrix(~Sex -1, cats)
colnames(treatment) <- c("Unirradiated", "Gamma", "HZE")
cats <- data.frame(cats, treatment, sex)

# format relevant data as a list
data <- list(CAT = cats$Cataracts, Gamma = cats$Gamma, 
             HZE = cats$HZE, Male = cats$SexM, Family = cats$Family, 
             nFam = length(unique(cats$Family)), N = nrow(cats))

nIter <- 60000
nChains <- 3
nThin <- 1
BurnIn <- 10000
nAdapt <- 1000
ests <- summary(mod)$coef[,1] # pull starting values from frequentist model
var <- as.numeric(as.data.frame(VarCorr(mod))$vcov)
inits <- list(list("tau" = var+0.2, "b0" = ests[1]+0.5, "b1" = ests[2]+0.5, "b2" = ests[3]+0.5,
                   "b3" = ests[4]+0.2, "b4" = ests[5]+0.2, "b5" = ests[6]+0.2),
              list("tau" = var-0.2, "b0" = ests[1]-0.5, "b1" = ests[2]-0.5, "b2" = ests[3]-0.5,
                   "b3" = ests[4]-0.2, "b4" = ests[5]-0.2, "b5" = ests[6]-0.2),
              list("tau" = var, "b0" = ests[1], "b1" = ests[2], "b2" = ests[3],
                   "b3" = ests[4], "b4" = ests[5], "b5" = ests[6]))

# -- Compile and run the model
params <- c("b0", "b1", "b2", "b3", "b4", "b5", "sigma2")
set.seed(556)
model.fit <- jags(data = data,
                  inits = inits,
                  parameters.to.save = params,
                  model.file = "cat.jag",
                  n.chains = nChains,
                  n.iter = nIter,
                  n.burnin = BurnIn,
                  n.thin = nThin)
```

```{r bayes_summary}
mcmc.model <- as.mcmc(model.fit)
#summary(mcmc.model)

# Convert posterior distributions to probabilities?
posts <- mcmc.model[[3]][,-7] 
posts <- exp(posts)/(1 + exp(posts))
phpds <- HPDinterval(posts)
posts <- data.frame(posts)

# Create table of posterior estimates
means <- apply(posts, 2, mean)
medians <- apply(posts, 2, median)
mode_fun <- function(x) {
  ux <- round(unique(x), digits = 3)
  return(ux[which.max(tabulate(match(x, ux)))])
}
modes <- apply(round(posts, 4), 2, mode_fun)
sds <- apply(posts, 2, sd)
glm_probs <- exp(ests) / (1 + exp(ests))
p_var <- exp(var) / (1 + exp(var))
bayes_tab <- round(data.frame(c(ps, p_var), means, medians, modes, sds, phpds), digits = 3)
rownames(bayes_tab) <- c("$\\beta_0$", "$\\beta_1$", "$\\beta_2$", "$\\beta_3$", "$\\beta_4$", "$\\beta_5$", "$\\sigma^2$")
```

```{r bayes_tab}
kbl(bayes_tab,
    caption = "Final Model Parameter Estimates (probabilities)",
    col.names = c("GLMM Est", "MCMC Mean", "MCMC Median", "MCMC Mode", "MCMC SD", "HPD Lower", "HPD Upper")) %>%
  kable_styling(latex_options = "hold_position") %>%
  kable_classic_2(full_width = F, html_font = "Cambria", position = "float_left")
```



```{r postdens}
colors <- c("Density" = "#5C88DAFF", "Mode" = "#84BD00FF", "HPD" = "#CC0C00FF")
p0 <- ggplot(posts, aes(x = b0)) +
  geom_density(aes(color = "Density", fill = "Density"), alpha = 0.5) +
  geom_vline(aes(xintercept = bayes_tab[1,6], color = "HPD"), linetype = "dashed") +
  geom_vline(aes(xintercept = bayes_tab[1,7], color = "HPD"), linetype = "dashed") +
  geom_vline(aes(xintercept = bayes_tab[1,4], color = "Mode")) +
  scale_x_continuous(expand = c(0, 0)) +
  theme_light() +
  theme(plot.title = element_text(size= 12)) + 
  labs(x = "Probability",
       title = "$\beta_0$")

p1 <- ggplot(posts, aes(x = b1)) +
  geom_density(aes(color = "Density", fill = "Density"), alpha = 0.5) +
  geom_vline(aes(xintercept = bayes_tab[2,6], color = "HPD"), linetype = "dashed") +
  geom_vline(aes(xintercept = bayes_tab[2,7], color = "HPD"), linetype = "dashed") +
  geom_vline(aes(xintercept = bayes_tab[2,4], color = "Mode")) +
  scale_x_continuous(expand = c(0, 0)) +
  theme_light() +
  theme(plot.title = element_text(size= 12)) + 
  labs(x = "Probability",
       title = "$\beta_1$")

p2 <- ggplot(posts, aes(x = b2)) +
  geom_density(aes(color = "Density", fill = "Density"), alpha = 0.5) +
  geom_vline(aes(xintercept = bayes_tab[3,6], color = "HPD"), linetype = "dashed") +
  geom_vline(aes(xintercept = bayes_tab[3,7], color = "HPD"), linetype = "dashed") +
  geom_vline(aes(xintercept = bayes_tab[3,4], color = "Mode")) +
  scale_x_continuous(expand = c(0, 0)) +
  theme_light() +
  theme(plot.title = element_text(size= 12)) + 
  labs(x = "Probability",
       title = "$\beta_2$")

p3 <- ggplot(posts, aes(x = b3)) +
  geom_density(aes(color = "Density", fill = "Density"), alpha = 0.5) +
  geom_vline(aes(xintercept = bayes_tab[4,6], color = "HPD"), linetype = "dashed") +
  geom_vline(aes(xintercept = bayes_tab[4,7], color = "HPD"), linetype = "dashed") +
  geom_vline(aes(xintercept = bayes_tab[4,4], color = "Mode")) +
  scale_x_continuous(expand = c(0, 0)) +
  theme_light() +
  theme(plot.title = element_text(size= 12)) + 
  labs(x = "Probability",
       title = "$\beta_3$")

p4 <- ggplot(posts, aes(x = b4)) +
  geom_density(aes(color = "Density", fill = "Density"), alpha = 0.5) +
  geom_vline(aes(xintercept = bayes_tab[5,6], color = "HPD"), linetype = "dashed") +
  geom_vline(aes(xintercept = bayes_tab[5,7], color = "HPD"), linetype = "dashed") +
  geom_vline(aes(xintercept = bayes_tab[5,4], color = "Mode")) +
  scale_x_continuous(expand = c(0, 0)) +
  theme_light() +
  theme(plot.title = element_text(size= 12)) + 
  labs(x = "Probability",
       title = "$\beta_4")

p5 <- ggplot(posts, aes(x = b5)) +
  geom_density(aes(color = "Density", fill = "Density"), alpha = 0.5) +
  geom_vline(aes(xintercept = bayes_tab[6,6], color = "HPD"), linetype = "dashed") +
  geom_vline(aes(xintercept = bayes_tab[6,7], color = "HPD"), linetype = "dashed") +
  geom_vline(aes(xintercept = bayes_tab[6,4], color = "Mode")) +
  scale_x_continuous(expand = c(0, 0)) +
  theme_light() +
  theme(plot.title = element_text(size= 12)) + 
  labs(x = "Probability",
       title = "$\beta_5$")


p6 <- ggplot(posts, aes(x = sigma2)) +
  geom_density(aes(color = "Density", fill = "Density"), alpha = 0.5) +
  geom_vline(aes(xintercept = bayes_tab[7,6], color = "HPD"), linetype = "dashed") +
  geom_vline(aes(xintercept = bayes_tab[7,7], color = "HPD"), linetype = "dashed") +
  geom_vline(aes(xintercept = bayes_tab[7,4], color = "Mode")) +
  scale_x_continuous(expand = c(0, 0)) +
  theme_light() +
  theme(plot.title = element_text(size= 12)) + 
  labs(x = "Probability",
       title = "$\\sigma^2$")

grid.arrange(p0, p1, p2, p3, p4, p5, p6, 
             ncol = 3, top = "Posterior Densities with 95% HPD Intervals")
```

## Limitations and Alternatives

Presence of competing risks in the experimental design obscure the presence of family clusters of cataracts. It is reasonable to assume that individuals in the irradiated Treatment groups were more likely to die due to more acute effects of irradiation before developing cataracts, compared to individuals in the control group. Simplifying the dataset by instituting a cutoff age attempts to address this limitation, but the issue persists even after excluding individuals deceased before Age = 552.

## Results and Conclusions  


The final probability for developing cataracts can be written as:  
$$
p_i = 0.063 + 0.55*Gamma_i + 0.623*HZE_i + 0.708*Male_i + 0.726*Gamma_i*Male_i + 0.770*HZE_i*Male_i
$$

```{r contr}
cats_emms <- emmeans(mod, ~ Treatment | Sex, infer = TRUE, type = "response")
cats_emms
pairs(cats_emms, reverse = TRUE)
emmip(cats_emms, Treatment ~ Sex) + theme_light() + 
  ggtitle("Predicted Probabilities of Cataracts \nby Sex and Treatment Group") + scale_color_startrek()
```

Assessing odds ratios for cataracts between groups:  

```{r oddsr}
cats_emms <- emmeans(mod, ~ Treatment | Sex, infer = TRUE, type = "response")
podds <- pairs(cats_emms, reverse = TRUE)
podds <- as.data.frame(confint(podds)) %>%
  rename(Contrast = contrast, Lower = asymp.LCL, Upper = asymp.UCL)

or1_plot <- ggplot(podds, aes(x = odds.ratio, y = Contrast, xmin = Lower, xmax = Upper)) +
  geom_errorbarh(aes(height = 0.2, color = Sex),
                 position = position_dodge(0.3), lwd = 1) +
  geom_point(aes(color = Sex), position = position_dodge(0.3)) +
  geom_text(aes(label = round(odds.ratio, 2), color = Sex), position = position_dodge(0.7)) + 
  geom_vline(aes(xintercept = 1), color = "#84BD00FF", lty = 2) +
  theme_light() +
  theme(axis.text.y = element_text(angle = 45, vjust = 1, hjust = 0.5)) +
  scale_color_startrek() +
  labs(x = "Odds Ratio",
       title = "Odds Ratios of Developing Cataracts \nby Treatment Group within Sex")

or1_plot
```

```{r oddsr2}
cats_emms1 <- emmeans(mod, ~Sex | Treatment, infer = TRUE, type = "response")
podds1 <- pairs(cats_emms1, reverse = TRUE)
podds1
podds1 <- as.data.frame(confint(podds1)) %>%
  rename(Contrast = contrast, Lower = asymp.LCL, Upper = asymp.UCL)

or2_plot <- ggplot(podds1, aes(x = odds.ratio, y = Contrast, xmin = Lower, xmax = Upper)) +
  geom_errorbarh(aes(height = 0.2, color = Treatment),
                 position = position_dodge(0.3), lwd = 1) +
  geom_point(aes(color = Treatment), position = position_dodge(0.3)) +
  geom_text(aes(label = round(odds.ratio, 2), color = Treatment), 
            position = position_dodge(0.3), vjust = -1) +
  theme_light() +
  geom_vline(aes(xintercept = 1), color = "#FFCD00FF", lty = 2) +
  scale_color_startrek() +
  labs(x = "Odds Ratio",
       title = "Odds Ratio of Developing \nCataracts by Sex within Treatment")

or2_plot
```  

Assessing relative risk for cataracts between groups:  
```{r RR}
em1 <- emmeans(mod, ~ Treatment|Sex)
em1log <- regrid(em1, "log")
rrs1 <- contrast(em1log, interaction = "revpairwise", type = "response")
rrs1 <- as.data.frame(confint(rrs1)) %>%
  rename(Contrast = Treatment_revpairwise, Lower = asymp.LCL, Upper = asymp.UCL)

rr1_plot <- ggplot(rrs1, aes(x = ratio, y = Contrast, xmin = Lower, xmax = Upper)) +
  geom_errorbarh(aes(height = 0.2, color = Sex),
                 position = position_dodge(0.3), lwd = 1) +
  geom_point(aes(color = Sex), position = position_dodge(0.3)) +
  geom_text(aes(label = round(ratio, 2), color = Sex), 
            position = position_dodge(0.7)) +
  theme_light() +
  geom_vline(aes(xintercept = 1), color = "#84BD00FF", lty = 2) +
  scale_color_startrek() +
  labs(x = "Relative Risk",
       title = "Relative Risk of Developing \nCataracts by Sex within Treatment")

rr1_plot
```
  
```{r RR1}
em2 <- emmeans(mod, ~ Sex|Treatment)
em2log <- regrid(em2, "log")
rrs2 <- contrast(em2log, interaction = "revpairwise", type = "response")
rrs2 <- as.data.frame(confint(rrs2)) %>%
  rename(Contrast = Sex_revpairwise, Lower = asymp.LCL, Upper = asymp.UCL)

rr2_plot <- ggplot(rrs2, aes(x = ratio, y = Contrast, xmin = Lower, xmax = Upper)) +
  geom_errorbarh(aes(height = 0.2, color = Treatment),
                 position = position_dodge(0.3), lwd = 1) +
  geom_point(aes(color = Treatment), position = position_dodge(0.3)) +
  geom_text(aes(label = round(ratio, 2), color = Treatment), 
            position = position_dodge(0.3), vjust = -1) +
  theme_light() +
  geom_vline(aes(xintercept = 1), color = "#FFCD00FF", lty = 2) +
  scale_color_startrek() +
  labs(x = "Relative Risk",
       title = "Relative Risk of Developing \nCataracts by Treatment within Sex")

rr2_plot
```

Visualizing random effect by family:  
```{r re_plot}
p_sigs <- bayes_tab[7,]
est <- as.numeric(p_sigs[1])
psig <- as.numeric(p_sigs[4])
hpdl <- as.numeric(p_sigs[6])
hpdu <- as.numeric(p_sigs[7])
REs <- augment(ranef(mod,condVar = TRUE), ci.level = 0.95) %>%
  select(c(level, estimate, lb, ub)) %>%
  rename(Family = level) %>%
  mutate(Prob = exp(estimate)/(1+exp(estimate)),
         Lower = exp(lb)/(1+exp(lb)),
         Upper = exp(ub)/(1+exp(ub)))
colors <- c("Family Effect" = "#5C88DAFF", "GLMM Est" = "#CC0C00FF", "Bayes Mode" = "#84BD00FF",
            "HPD Interval" = "#FFCD00FF")
re_plot <- ggplot(REs, aes(x = Prob, y = Family, xmin = Lower, xmax = Upper)) +
  geom_errorbarh(aes(height = 0, color = "Family Effect")) +
  geom_point(aes(color = "Family Effect")) +
  geom_vline(aes(xintercept = est, color = "GLMM Est"), lwd = 1, lty = 2) +
  geom_vline(aes(xintercept = psig, color = "Bayes Mode"), lwd = 1, lty = 2) +
  geom_vline(aes(xintercept = hpdl, color = "HPD Interval"), lwd = 1, lty = 4) +
  geom_vline(aes(xintercept = hpdu, color = "HPD Interval"), lwd = 1, lty = 4) +
  theme_light() +
  scale_color_manual(values = colors) +
  labs(color = "", title = "Random Effect by Family")

re_plot
```

Female mice are less likely to develop cataracts than male mice across all treatment groups. Treatment is strongly associated with differences in the probability of male mice developing cataracts, but it is not strongly associated with differences in the probability of female mice developing cataracts.  
Female mice have a lower probability of developing cataracts than male mice across all 3 treatment groups. The difference in probability of developing cataracts for females is very small and doesn't show a strong association. However, the differences in probability of developing cataracts among males in the three treatment groups are quite large and do indicate a strong association.

GLMM model selection and Bayesian HPD intervals indicate that Family is an important source of variation in predicting cataract development.  

## Authors' Statements

Both authors contributed to statistical analysis and writing for this report. Both authors read and approved the final report. Both authors contributed equally to introduction, exploratory data analysis, summary statistics, and results. Alyssa Allsop served as the primary statistician on fitting Frequentist models; Amira Burns served as the primary statistician on fitting the Bayesian model.

## References

Agresti, A (2013). *Categorical Data Analysis* (Third Edition). John Wiley & Sons.

Chernyavskiy, P., Edmondson, E. F., Weil, M. M., & Little, M. P. (2017). *High-energy particle beam and gamma radiation exposure, familial relatedness and cancer in mice.* British journal of cancer, 117(1), 41--50. <https://doi.org/10.1038/bjc.2017.141>

Gelman, A., Carlin, J., Stern, H., Dunson, D., Vehtari, A., Rubin, D. (2014). *Bayesian Data Analysis* (Third Edition). CRC Press.

Roback, P., Legler, J. (2021). *Beyond Multiple Linear Regression. Applied Generalized Linear Models and Multilevel Models in R* (Second Edition). CRC Press. Online version: <https://bookdown.org/roback/bookdown-BeyondMLR/>

DiMaggio, C., Albert, J., Spiegelhalter, D., Best, N., Gelman, A., Carstensen, B., Guerrin, L., Jensen, S. (2015) *Bayesian Analysis for Epidemiologists III: Regression Modeling.* Center for Injury Epidemiology and Prevention at Columbia University. <http://www.columbia.edu/~cjd11/charles_dimaggio/DIRE/styled-4/styled-11/code-8/#bayesian-regression-modeling-with-bugs-or-jags>  
  
R Packages:  
 + Douglas Bates, Martin Maechler, Ben Bolker, Steve Walker (2015). Fitting Linear Mixed-Effects Models Using lme4. Journal of Statistical Software, 67(1), 1-48. doi:10.18637/jss.v067.i01.  
 + Martyn Plummer (2021). rjags: Bayesian Graphical Models using MCMC. R package version 4-12. https://CRAN.R-project.org/package=rjags  
 + Russell V. Lenth (2022). emmeans: Estimated Marginal Means, aka Least-Squares Means. R package version 1.7.3. https://CRAN.R-project.org/package=emmeans  

## Appendix  

### Additional Exploratory Data Analysis

### GLMM Model Selection

```{r aic}
#AIC comparison table
row1 <- c("Sex", "Unirradiated", "HZE", "Gamma")
Male <- c( 0.14, 0.47, 0.35)
Female <- c( 0.06, 0.10, 0.08)
dt <- rbind(Male, Female)
colnames(dt) <- c("Unirradiated", "HZE", "Gamma")

aics <- c(round(AIC(final_mod),2), round(AIC(full_mod),2), round(AIC(fixed_mod),2), round(AIC(no_interaction),2))
models <- c("Final Model", "Full Mixed Model", "Fixed Effects Model", "Mixed Model with no Interaction")
dt <- cbind(models,aics)
colnames(dt) <- c("Model", "AIC")

dt %>% 
  kbl(caption = "AIC Comparison Across Different Models") %>% 
  kable_classic_2(full_width = T, html_font = "Cambria", )
```

### GLMM Model Diagnostics (necessary?)

### Bayes Model Diagnostics  
Traceplots:
```{r trace}
params = c("b0", "b1", "b2", "b3", "b4", "b5", "sigma^2")
traplot(mcmc.model, parms = params )
```  
  
ACF plots:  
```{r acf}
par(mfrow = c(2,2))
acf(mcmc.model[[3]][,1], main = "b0")
acf(mcmc.model[[3]][,2], main = "b1")
acf(mcmc.model[[3]][,3], main = "b2")
acf(mcmc.model[[3]][,4], main = "b3")
par(mfrow = c(2,2))
acf(mcmc.model[[3]][,5], main = "b4")
acf(mcmc.model[[3]][,6], main = "b5")
acf(mcmc.model[[3]][,8], main = "sigma2")
```
Gelman-Rubin Diagnostics:  
```{r grd}
gelman.diag(mcmc.model)
gelman.plot(mcmc.model)
```


### Post-Hoc Assessments

```{r glmm_ors}
plot_model(mod, sort.est = TRUE, show.values = TRUE, type = "int",
           color = "Dark2", vline.color = "darkorchid3",
           width = 0.1, title = "Final Model: Probabilities of Cataracts by Sex, Treatment Group")
tab_model(mod, show.re.var = TRUE,
          pred.labels = c("Control(Female)", "Gamma(Female)", "HZE(Female)",
                          "Control(Male)", "Gamma(Male)", "HZE(Male)"),
          dv.labels = "Final Model: Odds Ratios of Cataracts by Sex, Treatment Group")
```

### Full Code  
```{r code, include=FALSE, echo = TRUE}


```

