---
title: "Three (Groups of) Blind Mice" 
subtitle: "Familial Clusters of Cataract Development in Irradiated Mice"
author: "Alyssa Allsop and Amira Burns"
date: "`r Sys.Date()`"
output: 
  ioslides_presentation: 
    widescreen: true
    fontsize: 10
---

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```  

```{r, include=FALSE}

library(readxl)
library(tidyverse)
library(lme4)
library(broom.mixed)
library(kableExtra)
library(ggsci)
library(coda)
library(rjags)
library(R2jags)
library(superdiag)
library(mcmcplots)
library(ggmcmc)
library(emmeans)
```

  

## Introduction
  
- Research Questions:   
    + Are mice in some families more likely to develop more severe cataracts?
- Experimental Design:  
    + 1820 genetically heterogeneous mice from 48 unique families were randomly assigned, with equal family weights, to three treatment groups: Gamma radiation, HZE radiation, and unirradiated control. The simplified dataset records symptoms at time of death for mice of Age $\ge$ 552 days, for a total sample size of 1169.   
    + **Response**: Cataracts Status, either 0 or 1. Where 1 indicates the presence of cataracts that are of clinical concern.  
    + **Primary predictors under consideration**: Treatment, Family  
    + **Covariates**: Sex, Weight, Coat Color, Body Condition Score, Age, Myeloid Leukemia, Harderian Tumor, PreT Lymphoma


## Summary Tables & Plots {.smaller}
```{r load_data}
cats <- read_excel("GRSD.cataract.xlsx", sheet = "Sheet1")
```

```{r format_data, include = FALSE}
# remove spaces from column and value names
names(cats) <- str_replace_all(names(cats), " ", "_")
cats <- cats %>%
  mutate(CoatColor = str_replace_all(coat_color, " ", "_"))
str(cats)

# turn categorical vars into factor
cats <- cats %>%
  rename(Age = `age_(days)`,
         Weight = weight,
         Animal = animal) %>%
  mutate(Sex = as.factor(sex),
         CoatColor = as.factor(CoatColor),
         Family = as.factor(family), # should this stay a factor? Yes?
         BCS = as.ordered(BCS),
         Treatment = relevel(as.factor(groups), ref = "Unirradiated"),
         MyeloidLeukemia = as.factor(Myeloid_Leukemia),
         HarderianTumor = as.factor(Harderian_Tumor),
         PreTLymphoma = as.factor(PreT_Lymphoma),
         Score = as.ordered(Cataract_Score)) # ordinal cat; leave as numeric?

# select vars, add binary conversion for score
cats <- cats %>%
  select(c(Animal, Sex, Weight, CoatColor, Family, BCS, Age, Treatment,
           MyeloidLeukemia, HarderianTumor, PreTLymphoma, Score)) %>%
  mutate(Cataracts = ifelse(Score < 2, 0, 1))

# glance at the dataset
# str(cats)

sex_trt <- cats %>%
  group_by(Treatment, Sex, Cataracts) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  group_by(Treatment, Sex) %>%
  mutate(prop = round(n/sum(n), digits = 2))

sum_tab <- sex_trt %>%
  pivot_wider(names_from = Sex, values_from = c(n, prop)) %>%
  mutate(Cataracts = ifelse(Cataracts == 0, "No", "Yes"),
         Treatment = rep(" ", 1)) %>%
  relocate(prop_F, .after = n_F)
n_C <- c(438, 58, 214, 63, 281, 115)
prop_C <- c(0.88, 0.12, 0.77, 0.23, 0.71, 0.29)

sum_tab <- sum_tab %>%
  ungroup() %>%
  mutate(n_C = n_C, prop_C = prop_C) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 258, prop_F = NA,
          n_M = 238, prop_M = NA, n_C = 496, prop_C = NA, .after = 2) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 145, prop_F = NA,
          n_M = 132, prop_M = NA, n_C = 277, prop_C = NA, .after = 5) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 205, prop_F = NA,
          n_M = 191, prop_M = NA, n_C = 396, prop_C = NA, .after = 8) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 608, prop_F = NA,
          n_M = 561, prop_M = NA, n_C = 1169, prop_C = NA, .after = 9)

grsex_score <- cats %>%
  group_by(Sex, Treatment, Family) %>%
  summarize(mean_score = mean(Cataracts))
```

```{r}
ggplot(grsex_score, aes(x = Treatment, y = mean_score, color = Sex)) +
  geom_line(aes(group = Family)) + facet_wrap(vars(Sex)) +
  scale_color_startrek() +
  scale_x_discrete(expand = c(0, .2)) +
  theme_light() +
  labs(y = "mean score",
       title = "Cataracts by Family, Sex, Treatment Group")

```

<<<<<<< HEAD

## Mixed Logistic Regression
\[
\begin{aligned}
Y \sim\ &Binomial(1169, p)\ \ \mbox{or parameterized as   }\  Y_{ij} \sim Bernoulli(p) \\
log(\frac{p}{1-p}) = &\ \beta_0\ +\beta_1*Gamma_i*F_i\ + \beta_2*HZE_i*F_i\ + \\ &\beta_3*Unirradiated_i*M_i\ +\beta_4*Gamma_i*M_i\ + \beta_5*HZE_i*M_i\ + \\
[&v_{j} + \epsilon_{ij}] \\
&i = 1, ..., 1169\ \mbox{ mice} \\
&j = 1,...,47\ \ \mbox{ families}
\end{aligned}
\]  
=======
```{r glmm, include = FALSE}
# mixed model binomial logistic regression using adaptive Gauss-Hermite quadrature (nAGQ = 25)
mixed_mod <- glmer(Cataracts ~ Sex*Treatment + (1|Family), nAGQ = 25, data = cats, family = binomial)
```
>>>>>>> 7023ab25d3cbf9693dea3560024c7e3d7f8ac8f6



## Final Model {.smaller}
\vspace{}
- Frequentist approach
\vspace{}
$Y_{ij} \sim Bernoulli(p)\\log(\frac{p}{1-p}) = \beta_0 + \beta_1*Unirradiated_i*F_i + \beta_2*Gamma_i*F_i +  \beta_3*HZE*F_i + 
\\\beta_4*Unirradiated_i*M_i +  \beta_5*Gamma_i*M_i +  \beta_6*HZE_i*M_i +
[v_{j} + \epsilon_{ij}] \\ i = 1,...,1169 \: \mbox{mice} \\ j = 1,...,47 \: \mbox{families} \\ \mbox{and} \: v_j \sim N(0, \tau)$
\vspace{}
- Bayesian approach with non-informative priors
\vspace{}
\[
\begin{aligned}
\beta_0 &\sim N(0, 0.001)\\
\beta_1 &\sim N(0, 0.001)\\
\beta_2 &\sim N(0, 0.001)\\
\beta_3 &\sim N(0, 0.001)\\
\beta_4 &\sim N(0, 0.001)\\
\beta_5 &\sim N(0, 0.001)\\
\tau &\sim Gamma(0.001, 0.001) \mbox{ where}\ \sigma^2 = 1/\tau\\
\end{aligned}
\]


  
## Results  

```{r, include = FALSE}
cats_emms <- emmeans(mixed_mod, ~ Treatment | Sex, infer = TRUE, type = "response")
pairs(cats_emms, reverse = TRUE)
```

```{r}
emmip(cats_emms, Treatment ~ Sex) + theme_light() + ggtitle("Predicted Probabilities of Cataracts \nby Sex and Treatment Group") + scale_color_startrek() + ylab("Predicted Probabilities")
```
  

## Conclusions
- Our frequentist model shows the difference in cataract development among treatment groups is much larger among male rats
- Overall, the HZE group had the highest development of cataracts, followed by the Gamma group and then the unirradiated rats.
- Limitations?  

## References {.smaller}
Agresti, A (2013). *Categorical Data Analysis* (Third Edition). John Wiley & Sons.  
  
Chernyavskiy, P., Edmondson, E. F., Weil, M. M., & Little, M. P. (2017). *High-energy particle beam and gamma radiation exposure, familial relatedness and cancer in mice.* British journal of cancer, 117(1), 41–50. https://doi.org/10.1038/bjc.2017.141  
  
Gelman, A., Carlin, J., Stern, H., Dunson, D., Vehtari, A., Rubin, D. (2014). *Bayesian Data Analysis* (Third Edition). CRC Press.  
  
Roback, P., Legler, J. (2021). *Beyond Multiple Linear Regression. Applied Generalized Linear Models and Multilevel Models in R* (Second Edition). CRC Press. Online version: https://bookdown.org/roback/bookdown-BeyondMLR/  
  
DiMaggio, C., Albert, J., Spiegelhalter, D., Best, N., Gelman, A., Carstensen, B., Guerrin, L., Jensen, S. (2015) *Bayesian Analysis for Epidemiologists III: Regression Modeling.* Center for Injury Epidemiology and Prevention at Columbia University.  http://www.columbia.edu/~cjd11/charles_dimaggio/DIRE/styled-4/styled-11/code-8/#bayesian-regression-modeling-with-bugs-or-jags

  
# Any Questions?  
