---
title: "Three (Groups of) Blind Mice" 
subtitle: "Familial Clusters of Cataract Development in Irradiated Mice"
author: "Alyssa Allsop and Amira Burns"
date: "`r Sys.Date()`"
output: 
  ioslides_presentation: 
    widescreen: true
    fontsize: 10
---

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```  

```{r, include=FALSE}

library(readxl)
library(tidyverse)
library(lme4)
library(broom.mixed)
library(kableExtra)
library(ggsci)
library(coda)
library(rjags)
library(R2jags)
library(superdiag)
library(mcmcplots)
library(ggmcmc)
```

  

## Introduction
  
- Research Questions:   
    + Is there a genetic susceptibility to cataracts caused by radiation?  
    + Accounting for potential genetic susceptibility, is there a difference in cataract presentation between HZE radiation and gamma radiation? 
- Experimental Design:  
    + 1820 genetically heterogeneous mice from 48 unique families were randomly assigned, with equal family weights, to three treatment groups: Gamma radiation, HZE radiation, and unirradiated control. The simplified dataset records symptoms at time of death for mice of Age $\ge$ 552 days, for a total sample size of 1169.   
    + **Response**: Cataracts score, ordinal categorical variable on a scale 0-5. Score $\ge$ 2 indicates presence of cataracts; because sample sizes in ordinal categories are small, we decided to convert the response into a binary variable indicating absence or presence of cataracts.  
    + **Primary predictors under consideration**: Treatment, Family  
    + **Covariates**: Sex, Weight, Coat Color, Body Condition Score, Age, Myeloid Leukemia, Harderian Tumor, PreT Lymphoma
- Methods Overview:  
    + Mixed Logistic Regression with Treatment as a fixed effect, Family as a random effect, and accounting for any practically significant covariates. Obtain estimates for odds ratios. 
    + Bayesian Mixed Logistic Regression counterpart to Frequentist final model to obtain probability distributions for odds ratios.  


## Summary Tables & Plots {.smaller}
```{r load_data}
cats <- read_excel("GRSD.cataract.xlsx", sheet = "Sheet1")
```

```{r format_data, include = FALSE}
# remove spaces from column and value names
names(cats) <- str_replace_all(names(cats), " ", "_")
cats <- cats %>%
  mutate(CoatColor = str_replace_all(coat_color, " ", "_"))
str(cats)

# turn categorical vars into factor
cats <- cats %>%
  rename(Age = `age_(days)`,
         Weight = weight,
         Animal = animal) %>%
  mutate(Sex = as.factor(sex),
         CoatColor = as.factor(CoatColor),
         Family = as.factor(family), # should this stay a factor? Yes?
         BCS = as.ordered(BCS),
         Treatment = relevel(as.factor(groups), ref = "Unirradiated"),
         MyeloidLeukemia = as.factor(Myeloid_Leukemia),
         HarderianTumor = as.factor(Harderian_Tumor),
         PreTLymphoma = as.factor(PreT_Lymphoma),
         Score = as.ordered(Cataract_Score)) # ordinal cat; leave as numeric?

# select vars, add binary conversion for score
cats <- cats %>%
  select(c(Animal, Sex, Weight, CoatColor, Family, BCS, Age, Treatment,
           MyeloidLeukemia, HarderianTumor, PreTLymphoma, Score)) %>%
  mutate(Cataracts = ifelse(Score < 2, 0, 1))

# glance at the dataset
# str(cats)

sex_trt <- cats %>%
  group_by(Treatment, Sex, Cataracts) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  group_by(Treatment, Sex) %>%
  mutate(prop = round(n/sum(n), digits = 2))

sum_tab <- sex_trt %>%
  pivot_wider(names_from = Sex, values_from = c(n, prop)) %>%
  mutate(Cataracts = ifelse(Cataracts == 0, "No", "Yes"),
         Treatment = rep(" ", 1)) %>%
  relocate(prop_F, .after = n_F)
n_C <- c(438, 58, 214, 63, 281, 115)
prop_C <- c(0.88, 0.12, 0.77, 0.23, 0.71, 0.29)

sum_tab <- sum_tab %>%
  ungroup() %>%
  mutate(n_C = n_C, prop_C = prop_C) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 258, prop_F = NA,
          n_M = 238, prop_M = NA, n_C = 496, prop_C = NA, .after = 2) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 145, prop_F = NA,
          n_M = 132, prop_M = NA, n_C = 277, prop_C = NA, .after = 5) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 205, prop_F = NA,
          n_M = 191, prop_M = NA, n_C = 396, prop_C = NA, .after = 8) %>%
  add_row(Treatment = " ", Cataracts = "$\\sum$", n_F = 608, prop_F = NA,
          n_M = 561, prop_M = NA, n_C = 1169, prop_C = NA, .after = 9)

grsex_score <- cats %>%
  group_by(Sex, Treatment, Family) %>%
  summarize(mean_score = mean(Cataracts))
```

```{r}
ggplot(grsex_score, aes(x = Treatment, y = mean_score, color = Sex)) +
  geom_line(aes(group = Family)) + facet_wrap(vars(Sex)) +
  scale_color_startrek() +
  scale_x_discrete(expand = c(0, .2)) +
  theme_light() +
  labs(y = "mean score",
       title = "Cataracts by Family, Sex, Treatment Group")

```


## Mixed Logistic Regression
<div style="float: left; width: 40%;">
* Predictors Included in the Final Model:
  + Sex
  + Treatment group (Control, Gamma, or HZE)
  + Random intercept for family
</div>

<div style="float: right; width: 60%;">
* Other Predictors Considered:
  + weight
  + coat color
  + BCS (body condition score)
  + harderian tumor (bilateral, unilateral, none)
  + Myeloid leukemia,
  + PreT lymphoma
</div>

```{r glmm, include = FALSE}
# mixed model binomial logistic regression using adaptive Gauss-Hermite quadrature (nAGQ = 25)
mix_mod_full <- glmer(Cataracts ~ Sex + Weight + CoatColor + BCS + HarderianTumor + 
                        MyeloidLeukemia + PreTLymphoma + Treatment + (1|Family), nAGQ = 25, 
                      data = cats, family = binomial)
mixed_mod <- glmer(Cataracts ~ Sex*Treatment + (1|Family), nAGQ = 25, data = cats, family = binomial)
fixed_mod_full <- glm(Cataracts ~ Sex + Weight + CoatColor + BCS + HarderianTumor + 
                        MyeloidLeukemia + PreTLymphoma + Treatment, data = cats, family = binomial)
fixed_mod <- glm(Cataracts ~ Treatment*Sex, data = cats, family = binomial)


# test full vs reduced mixed models
anova(mix_mod_full, mixed_mod)

summary(mix_mod_full)
summary(mixed_mod)
summary(fixed_mod)
```

```{r}
#table of AIC values


aics <- c(AIC(mix_mod_full), AIC(mixed_mod), AIC(fixed_mod))
bics <- c(BIC(mix_mod_full), BIC(mixed_mod), BIC(fixed_mod))
dt <- rbind(aics, bics)
colnames(dt) <- c("Full Mixed Model", "Reduced Mixed Model", "Fixed Effects Model")
rownames(dt) <- c("AIC", "BIC")

dt %>% 
  kbl(caption = "AIC and BIC Comparisons Across Models", digits = 2) %>% 
  kable_classic_2(full_width = T, html_font = "Cambria")

```

## Mixed Logistic Regression {.smaller}

* The final model can be written as:  
\[
\begin{aligned}
Y \sim\ &Binomial(1169, p)\ \ \mbox{or parameterized as   }\  Y_{ij} \sim Bernoulli(p) \\
log(\frac{p}{1-p}) = &\ \beta_0*Unirradiated_i*F_i\ +\beta_1*Gamma_i*F_i\ + \beta_2*HZE_i*F_i\ + \\ &\beta_3*Unirradiated_i*M_i\ +\beta_4*Gamma_i*M_i\ + \beta_5*HZE_i*M_i\ + \\
[&v_{j} + \epsilon_{ij}] \\
&i = 1, ..., 1169\ \mbox{ mice} \\
&j = 1,...,47\ \ \mbox{ families}
\end{aligned}
\]  



## Bayesian Mixed Logistic Regression
<div class='"columns-2">

The final Bayesian model can be written as:  
\[
\begin{aligned}
Y \sim Binomial&(1169, p) \ \mbox{   or parameterized as   }\ \ \  Y_{ij} \sim Bernoulli(p)\\
log(\frac{p}{1-p}) = &\ \beta_0*Control_i*F_i\ +\beta_1*Gamma_i*F_i\ + \beta_2*HZE_i*F_i\ + \\ &\beta_3*Control_i*M_i\ +\beta_4*Gamma*M\ + \beta_5*HZE*M\ + \\
&v_{ij} + \epsilon_{ij}\\
&i = 1, ..., 1169\ \mbox{ mice} \\
&j = 1,...,47\ \ \mbox{ families, and} \\
v_j \sim N(0, \tau)
\end{aligned}
\]  


With non-informative parameters:  
\[
\begin{aligned}
\beta_0 &\sim N(0, 0.001)\\
\beta_1 &\sim N(0, 0.001)\\
\beta_2 &\sim N(0, 0.001)\\
\beta_3 &\sim N(0, 0.001)\\
\beta_4 &\sim N(0, 0.001)\\
\beta_5 &\sim N(0, 0.001)\\
\tau &\sim Gamma(0.001, 0.001) \mbox{ where}\ \sigma^2 = 1/\tau\\
\end{aligned}
\]
 
</div>  
  
## Results & Conclusions  

- What can we infer about fixed and random effects from the models?


```{r}
#female control
#exp(-2.4865 -.2141)/(1 + exp(-2.4865 -.2141))
#female HZE
#exp(-2.4865 + .2879)/(1 + exp(-2.4865 + .2879))
#female in the gamma group is:
#exp(-2.4865)/(1 + exp(-2.4865))

# Male control
#exp(-2.4865 + 1.8647 -.2141 - .9753)/(1 + exp(-2.4865 + 1.8647 -.2141 - .9753))
#Male in HZE
#exp(-2.4865 + 1.8647 + .2879 + 0.2308)/(1 + exp(-2.4865 + 1.8647 + .2879 + 0.2308))
#Male Gamma
#exp(-2.4865 + 1.8647)/(1 + exp(-2.4865 + 1.8647))


row1 <- c("Sex", "Unirradiated", "HZE", "Gamma")
Male <- c( 0.14, 0.47, 0.35) ### need to update these with the new paramaterization of the model
Female <- c( 0.06, 0.10, 0.08)
dt <- rbind(Male, Female)
colnames(dt) <- c("Unirradiated", "HZE", "Gamma")


dt %>%
  kbl(caption = "Probabilities of Developing Cataracts of Clincal Concern") %>%
  kable_classic_2(full_width = T, html_font = "Cambria")

# dt %>%
#   kbl(caption = "Probabilities of Developing Cataracts of Clincal Concern") %>%
#   kable_minimal()
```
- Limitations?  
- What questions does this raise for future studies?  
  
## References  
  
# Any Questions?  
